# Activity 09/03/2021

## STAT 445 / 645 -- Section 1001 <br> Instructor: Colin Grudzien<br>

## Instructions:
We will work through the following series of activities as a group.  Follow the instructions in each sub-section when the instructor assigns you to a breakout room.

## Activities:

### Activity 1: Vectorization

#### Question 1:

Consider the R output of the matrix below:
```{r}
matrix(c(4, 1, 9, 5, 10, 7), ncol = 2, byrow = TRUE)
```
Examine each command below and try to figure out the output before typing them.

  1. `matrix(c(4, 1, 9, 5, 10, 7), nrow = 3)`
  2. `matrix(c(4, 9, 10, 1, 5, 7), ncol = 2, byrow = TRUE)`
  3. `matrix(c(4, 9, 10, 1, 5, 7), nrow = 2)`
  4. `matrix(c(4, 1, 9, 5, 10, 7), ncol = 2, byrow = TRUE)`



#### Question 2:
Given the following matrix:

```{r eval=FALSE}
m <- matrix(1:12, nrow=3, ncol=4)
m
```

What you think will happen when you run each of the following?

1. `m ^ -1`
2. `m * c(1, 0, -1)`
3. `m > c(0, 20)`
4. `m * c(1, 0, -1, 2)`

Did you get the output you expected? 



#### Question 3:

We're interested in looking at the sum of the
following sequence of fractions:

```{r, eval=FALSE}
 x = 1/(1^2) + 1/(2^2) + 1/(3^2) + ... + 1/(n^2)
```

This would be tedious to type out, and impossible for high values of
n.  Use vectorisation to compute x when n=100. What is the sum when
n=10,000?


#### Question 4:
Make a vector with the numbers 1 through 26.  Multiply the vector by 2, and give the resulting vector elements the names A through Z (hint: there is a built in vector called `LETTERS`).



### Activity 2: More on list subsetting

#### Question 1:
Given the following list:

```{r eval=FALSE}
xlist <- list(a = "Software Carpentry", b = 1:10, data = head(iris))
```

Using your knowledge of both list and vector subsetting, extract the number 2 from xlist.
Hint: the number 2 is contained within the "b" item in the list.

