# Activity 11/29/2021

## STAT 445 / 645 -- Section 1001 <br> Instructor: Colin Grudzien<br>

## Instructions:
We will work through the following series of activities as a group and hold small group work and discussions in Zoom Breakout Rooms.  Follow the instructions in each sub-section when the instructor assigns you to a breakout room.

## Activities:

```{r}
require("faraway")
```

### Activity 1: mixed effects of binary variables
We will note in the following how we can construct a mixed effect model from a binary variable.  The syntax in R is to use a `*` symbol instead of a `+` to introduce a mixed effect that can change the slope of the trend.  This refers mathematically to the following idea in a simple regression model:

Let $Z$ be the binary predictor that takes the values as follows:
$$\begin{align}
Z = 
\begin{cases}
1 & \text{ category 1} \\
0 & \text{ category 2}
\end{cases}
\end{align}$$
The variable $X$ will represent some standard, continuous scale predictor.

The resulting equation for the mixed effect model is
$$\begin{align}
Y = \beta_0 + \beta_1 X + \beta_2 Z + \beta_3 Z * X + \epsilon
\end{align}$$


#### Question 1:
Using this the above formula, derive how the mixed effect model changes the slope and intercept simultaneously when we shift between the two categories.

##### Solution:
Notice that if $Z=0$ we have the following model
$$\begin{align}
Y = \beta_0 + \beta_1 X + \epsilon.
\end{align}$$

On the other hand, if $Z=1$ we have the following model
$$\begin{align}
&Y = (\beta_0 +\beta_2)  + (\beta_1 + \beta_3) X  +\epsilon   \\
\Leftrightarrow & Y = \tilde{\beta}_0  + \tilde{\beta}_1 X + \epsilon
\end{align}$$
where we have now adjusted both the slope and intercept for the case of category 1, relative to category 0.


#### Question 2:

Construct the standard multiple regression model in which `gamble` is the response for all other variables as predictors in the `teengamb` data.  Then construct the mixed effect model with `sex * (status + income + verbal)`.  Examining the model summaries, what do you think about the mixed effects?  Can you construct a reduced size, mixed effect model that gets at the most important aspects?  Note, you should not remove the binary variable if there are significant mixed effects that depend on it.

##### Solution:

```{r}
lm_gamb <- lm(gamble ~ sex + status + income + verbal, teengamb)
lm_mixed <- lm(gamble ~ sex * (status + income + verbal), teengamb)
summary(lm_gamb)
summary(lm_mixed)
lm_mixed <- lm(gamble ~ sex * (income + verbal), teengamb)
summary(lm_mixed)
lm_gamb_mixed <- lm(gamble  ~ sex * income, teengamb)
summary(lm_gamb_mixed)
```


#### Question 3:
Using the final model in the last question, produce an F test to determine if we favor a mixed effect model versus the simple encoding of the binary adjustment to the intercept.

##### Solution:

```{r}
lm_gamb <- lm(gamble ~ sex + income, teengamb)
anova(lm_gamb,lm_gamb_mixed)
```

With the p-value on the order of $10^{-3}$ we favor the mixed effect model.

### Activity 2

#### Question 1:
Using the same mixed effect technique as in the last activity, try to systematically construct a mixed effect model in terms of `log(wage) ~ race * (educ + exper)`.  What do you notice?

##### Solution:

```{r}
lm_wages <- lm(log(wage) ~ educ + exper + race, uswages)
summary(lm_wages)
lm_wages_mixed <- lm(log(wage) ~ race * (educ + exper), uswages)
summary(lm_wages_mixed)
lm_wages_mixed <- lm(log(wage) ~ race * exper + educ, uswages)
summary(lm_wages_mixed)
```

If we remove the last insignificant mixed effect, we will arrive at the same model as before with the standard binary encoding.

#### Question 2:

Using an F test with the mixed effect model in `log(wage) ~ race * (educ + exper)`, what do you conclude about using the mixed effect here?


##### Solution:

```{r}
lm_wages_mixed <- lm(log(wage) ~ race * (educ + exper), uswages)
anova(lm_wages, lm_wages_mixed)
```

We favor the model in which there is no mixed effect.
