# Homework 4 -- 10/28/2020

## STAT 445 / 645 -- Section 1001 <br> Instructor: Colin Grudzien<br>

## Instructions:
We will work through the following series of activities as a group and hold small group work and discussions in Zoom Breakout Rooms.  Follow the instructions in each sub-section when the instructor assigns you to a breakout room.

The solutions of this activity will give you the answers for the Homework 4 assignment in Canvas. You will be given instructions in Canvas how to submit your answers to the below.  You may work with others on this homework assignment and consult outside resources, but everyone is responsible for their own submission in Canvas.

## Due: 11/03/2020 11:59 PM

## Activities:

```{r}
require("numDeriv")
```


We will code Newton's method in one variable in R.

Recall, Newton's method is given as follows:

* Step 0: Identify the objective function $f$ and the root value we are solving for, possibly redefining $\tilde{f}(x) = f(x) - b$
* Step 1: Set an initial guess `x_0`, a maximum number of iterations `max_iter` and an error tolerance `tol`
* Step 2: Compute each of $-\tilde{f}(x_0)$ and $\tilde{f}'(x_0)$
* Step 3: If $\tilde{f}'(x_0)=0$  (using a comparison the absolute value being very close to zero, e.g. comparing with `tol`) break the function and throw an error message, else define $x_1 = x_0 - \frac{\tilde{f}(x_0)}{\tilde{f}'(x_0)}$
* Step 4: If the number of iterations has exceeded `max_iter` or $\tilde{f}(x_1) \leq$`tol`, break the function and print the reason for completion, including the number of iterations spent.  Else, repeat steps 2 through 4 with $x_0 = x_1$.


#### Question 1:

In the below, write out your function so that it takes the arguments `x_0`, `max_iter` and `tol` as described above, such that it will return error and / or completion messages, along with the final value of `x_1` and the absolute difference of $\tilde{f}(x_1)$ from zero if the algorithm is successful.  Set default arguments for `max_iter = 100` and `tol=10e-14`.  Use the package `numDeriv` and the function `grad` to compute the derivative of $\tilde{f}$ at `x_0` when required.


#### Question 2:
We will define the polynomial function below:
```{r}
f <- function(x){
 return((x-4)*(x-1)*(x+3))  
}
```

Use your function with the following starting guesses and report the outputs:
```{r}
initial_conditions <- c(2.35287527, 2.35284172, 2.35283735, 2.352836327, 2.352836323)
```


Interpret what this says about the sensitivity of the method to an initial guess `x_0`.


#### Question 3:
We will define the polynomial function below:
```{r}
f <- function(x){
 return(x^3 - 2*x + 2)  
}
```

Use the following starting guesses and report the outputs:
```{r}
initial_conditions <- c(0)
```

Manually compute the locations of the first two iterations, $x_1,x_2$ defined from $x_0=0$.  How is this related to the output above?


#### Question 4:
We will define the polynomial function below:
```{r}
f <- function(x){
 return(1 - x^2)  
}
```

Use the following starting guess and report the outputs:
```{r}
initial_conditions <- c(0,0.5, -0.5)
```

