# Activity 09/01/2021

## STAT 445 / 645 -- Section 1001 <br> Instructor: Colin Grudzien<br>

## Instructions:
We will work through the following series of activities as a group.  Follow the instructions in each sub-section when the instructor assigns you to a breakout room.

## Activities:

### Activity 1: vectors and matrices


#### Question 1:
We can declare a matrix full of zeros:

```{r eval=FALSE}
matrix_example <- matrix(0, ncol=6, nrow=3)
matrix_example
```

And similar to other data structures, we can ask things about our matrix:

```{r eval=FALSE}
class(matrix_example)
typeof(matrix_example)
str(matrix_example)
dim(matrix_example)
nrow(matrix_example)
ncol(matrix_example)
```

What do you think will be the result of
`length(matrix_example)`?
Try it.
Were you right? Why / why not?


#### Question 2:

Consider the R output of the matrix below:
```{r}
matrix(c(4, 1, 9, 5, 10, 7), ncol = 2, byrow = TRUE)
```
Examine each command below and try to figure out the output before typing them.

  1. `matrix(c(4, 1, 9, 5, 10, 7), nrow = 3)`
  2. `matrix(c(4, 9, 10, 1, 5, 7), ncol = 2, byrow = TRUE)`
  3. `matrix(c(4, 9, 10, 1, 5, 7), nrow = 2)`
  4. `matrix(c(4, 1, 9, 5, 10, 7), ncol = 2, byrow = TRUE)`
  
### Activity 2: practice with subsetting data

#### Question 1:

Given the following code:
```{r}
x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
names(x) <- c('a', 'b', 'c', 'd', 'e')
print(x)
```

Come up with at least 2 different commands that will produce the following output:
```{r}
x[2:4]
```

#### Question 2:

Given the following code:

```{r eval=FALSE}
x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
names(x) <- c('a', 'b', 'c', 'd', 'e')
print(x)
```

Write a subsetting command to return the values in x that are greater than 4 and less than 7.

#### Question 3:

Selecting elements of a vector that match any of a list of components
is a very common data analysis task. For example, the gapminder data set
contains `country` and `continent` variables, but no information between
these two scales. Suppose we want to pull out information from southeast
Asia: how do we set up an operation to produce a logical vector that
is `TRUE` for all of the countries in southeast Asia and `FALSE` otherwise?

Suppose you have these data:
```{r}
seAsia <- c("Myanmar","Thailand","Cambodia","Vietnam","Laos")
require(gapminder)
## extract the `country` column from a data frame (we'll see this later);
## convert from a factor to a character;
## and get just the non-repeated elements
countries <- unique(as.character(gapminder$country))
print(countries)
```

There's a wrong way (using only `==`), which will give you a warning;
a clunky way (using the logical operators `==` and `|`); and
an elegant way (using `%in%`). See whether you can come up with all three
and explain how they (don't) work.


### Activity 3: more practice with multi-dimensional indices

#### Question 1:

Given the following code:
```{r}
m <- matrix(1:18, nrow=3, ncol=6)
print(m)
```

1. Which of the following commands will extract the values 11 and 14?

A. `m[2,4,2,5]`

B. `m[2:5]`

C. `m[4:5,2]`

D. `m[2,c(4,5)]`


#### Question 2:
Given the following list:

```{r eval=FALSE}
xlist <- list(a = "Software Carpentry", b = 1:10, data = head(iris))
xlist
```

Using your knowledge of both list and vector subsetting, extract the number 2 from xlist.
Hint: the number 2 is contained within the "b" item in the list.



