# Activity 10/27/2021

## STAT 445 / 645 -- Section 1001 <br> Instructor: Colin Grudzien<br>

## Instructions:
We will work through the following series of activities as a group and hold small group work and discussions in Zoom Breakout Rooms.  Follow the instructions in each sub-section when the instructor assigns you to a breakout room.

## Activities:
```{r}
require("numDeriv")
```

### Activity 1:

We will code Newton's method in one variable in R.

Recall, Newton's method is given as follows:

* Step 0: Identify the objective function $f$ and the root value we are solving for, possibly redefining $\tilde{f}(x) = f(x) - b$
* Step 1: Set an initial guess `x_0`, a maximum number of iterations `max_iter` and an error tolerance `tol`
* Step 2: Compute each of $-\tilde{f}(x_0)$ and $\tilde{f}'(x_0)$
* Step 3: If $\tilde{f}'(x_0)=0$  (using a comparison of this being very close to zero) break the function and throw an error message, else define $x_1 = x_0 - \frac{\tilde{f}(x_0)}{\tilde{f}'(x_0)}$
* Step 4: If the number of iterations has exceeded `max_iter` or $\tilde{f}(x_1) \leq$`tol`, break the function and print the reason for completion, including the number of iterations spent.  Else, repeat steps 2 through 4 with $x_0 = x_1$.


#### Question 1:

In the below, write out your function so that it takes the arguments `x_0`, `max_iter` and `tol` as described above, such that it will return error and / or completion messages, along with the final value of `x_1` and the absolute difference of $\tilde{f}(x_1)$ from zero if the algorithm is successful.  Set default arguments for `max_iter = 100` and `tol=10e-14`.  Use the package `numDeriv` and the function `grad` to compute the derivative of $\tilde{f}$ at `x_0` when required.

```{r}
newton <- function(fun, x_0, max_iter=100, tol=10e-14) {# step 1: load function, initial guess and parameters
   for (i in 1:max_iter){
       # step 2:  evaluate function at x_0 and evaluate the derivative at x_0
       f_0 <- fun(x_0) 
       f_prime_0 <- grad(fun, x_0) 
       if (abs(f_prime_0) <= tol){ 
          # step 3: check if the derivative is zero and if so, break with error message
          cat("Derivative at", x_0,"is equal to zero \n")
          break
       }
       # if not, we compute the iterated guess
       x_1 <- x_0 - (f_0 / f_prime_0)
       f_1 <- fun(x_1)
       
       # step 4: check if we have reached close enough to a zero,
       if (abs(f_1) <= tol) {
          #if true, return the root, the precision and the number of iterations
          cat("Root found at", x_1,"with precision", abs(f_1), "with", i, "iterations taken \n")
          break
       }
       else if (i == max_iter){
        # else if we reach the max number of iterations allowed, terminate with error message
        cat("Terminated at a maximum of", max_iter, "iterations \n")
          break
       }
       else {
        # finally, if no error was reached, and no root was found, repeat with a new initial guess at x_1
        x_0 <- x_1  
       }
  }
}
```

#### Question 2:
We will define the polynomial function below:
```{r}
f <- function(x){
 return((x-4)*(x-1)*(x+3))  
}
```

Use your function with the following starting guesses and report the outputs:
```{r}
initial_conditions <- c(2.35287527, 2.35284172, 2.35283735, 2.352836327, 2.352836323)
```


Interpret what this says about the sensitivity of the method to an initial guess `x_0`.

##### Solution:
```{r}
for (i in 1:length(initial_conditions)){
  newton(f, initial_conditions[i])
}
```

This says that there is extreme sensitivity to the choice of the initial condition in this example.  Very small differences lead to very different roots.

#### Question 3:
We will define the polynomial function below:
```{r}
f <- function(x){
 return(x^3 - 2*x + 2)  
}
```

Use the following starting guesses and report the outputs:
```{r}
initial_conditions <- c(0)
```

Manually compute the locations of the first two iterations, $x_1,x_2$ defined from $x_0=0$.  How is this related to the output above?

##### Solution:
```{r}
for (i in 1:length(initial_conditions)){
  newton(f, initial_conditions[i])
}
```

Note, we can compute
$$f'(x) = 3x - 2$$ 
and therefore
$$
x_1 = x_0 - \frac{f(x_0)}{f'(x_0)} = 0 - \frac{2}{-2} = 1
$$

Then computing
$$
x_2 = x_1 - \frac{f(x_1)}{f'(x_1)} = 1 - \frac{1}{1} =0
$$

Therefore, the algorithm will get stuck in an infinite loop.

#### Question 4:
We will define the polynomial function below:
```{r}
f <- function(x){
 return(1 - x^2)  
}
```

Use the following starting guess and report the outputs:
```{r}
initial_conditions <- c(0,0.5, -0.5)
```

##### Solution:
```{r}
for (i in 1:length(initial_conditions)){
  newton(f, initial_conditions[i])
}
```