# Activity 10/20/2021

## STAT 445 / 645 -- Section 1001 <br> Instructor: Colin Grudzien<br>

## Instructions:
We will work through the following series of activities as a group and hold small group work and discussions in Zoom Breakout Rooms.  Follow the instructions in each sub-section when the instructor assigns you to a breakout room.

## Activities:

To perform some matrix operations, it can be helpful to use some non-base functions designed for this.  We will require the library `matlab`

```{r}
require("matlab")
```



### Activity 1: Linear inverse problem

#### Question 1:
Suppose the matrix `A` from the last activity
```{r}
A <- matrix(seq(3, 27, by=3), nrow=3, ncol=3)
```
defines a linear inverse problem for an unknown vector $\mathbf{x}$ and an observed vector,
```{r}
b <- ones(3,1)
```

Is there any vector $\mathbf{x}$ that satisfies,
$$\mathbf{A} \mathbf{x} = \mathbf{b}?$$
Why does such a choice exist or why not?


##### Solution:
Notice that 
```{r, eval=FALSE}
solve(A,b)
```
gives a warning message about a singularity.  This is because of the zero eigenvalue that we discovered last time in this matrix.  Therefore, no formal inverse exists to this problem.


#### Question 2:

```{r}
b <- zeros(3,1)
b
```

Is there any vector $\mathbf{x}$ that satisfies,
$$\mathbf{A} \mathbf{x} = \mathbf{b}?$$
Why does such a choice exist or why not?

##### Solution:
Note that in this case, even though an inverse of `A` does not exist, the zero eigen value of `A` means that any scalar times this eigenvector is a solution:
```{r}
e_decomp <- eigen(A)
A %*% e_decomp$vectors[,3]
```


#### Question 3:
Set a random seed to be equal to zero and generate a 4-by-4 dimension matrix with entries given by a sample size $n=16$ from the standard normal.  Suppose that this defines the relationship between the observed variable
```{r}
b <- ones(4,1)
```
and an unknown vector $\mathbf{x}$ as,
$$\mathbf{A}\mathbf{x}= \mathbf{b}.$$

Does any `x` as above exist?  If one does find the solution.  Explain why the solution exists or not due to the eigenvalues of the matrix.


##### Solution:
```{r}
set.seed(0)
A <- matrix(rnorm(16), nrow=4, ncol=4)
det(A)
```
Notice, the determinant is non-zero and therefore the eigenvalues,
```{r}
eigen(A)
```
are all non-zero. Therefore an inverse exits and we can compute
```{r}
solve(A, b)
```


### Activity 2: matrix norms

Recall that matrix norms can be computed in R with the norm function.  In the following we will consider some of their properties.

#### Question 1:

Compute the eigenvalues of the matrix `B`  computed from matrix A below

```{r}
B <- t(A) %*% A
```

then compute the sum of the square values.

##### Solution:

For a symmetric matrix, the Frobenius norm is equal to the square root of the sum of square eigenvalues:

```{r}
eigen_decomp <- eigen(B)
vals <- eigen_decomp$values
sum(vals^(2))
```

#### Question 2:

Using the `norm` function, compute the norm-squared of the matrix `B`.  Try changing the default norm -- which choice of norm-squared looks familiar?

##### Solution

```{r}
norm(B, type="F")^2
```