# Activity 10/19/2020

## STAT 445 / 645 -- Section 1001 <br> Instructor: Colin Grudzien<br>

## Instructions:
We will work through the following series of activities as a group and hold small group work and discussions in Zoom Breakout Rooms.  Follow the instructions in each sub-section when the instructor assigns you to a breakout room.

## Activities:

To perform some matrix operations, it can be helpful to use some non-base functions designed for this.  We will require the library `matlab`

```{r}
require("matlab")
```


### Activity 1: Diagonalization

Suppose we are given the following matrix
```{r}
A <- matrix(seq(3, 27, by=3), nrow=3, ncol=3)
```

#### Question 1:
Compute the rank of the matrix `A` . Then compute the eigen values of the matrix `A`.  How is the rank related to the eigenvalues of the matrix?


#### Question 2:
Use the  eigen decomposition as follows: extract the matrix of the eigenvectors by calling the variable `$vectors` and assign these vectors to a matrix called `C`.  Check if it is possible to invert this matrix and if so, define the inverse `C_inv`.  Explain why this is possible or why not.



#### Question 3: 
Now try the following, multiply `A` by `C_inv` on the left and `C` on the right.  What do you notice about the values off diagonal? 

Try using the following comparison to see how close the elements are to the value zero,
```{r, eval=FALSE}
abs(product) <= ones(3) * 10e-14
```

Take the diagonal elements of the product above and find the absolute difference of these from the eigenvalues of `A`. What do you notice about the result.


### Activity 2: Linear inverse problem

#### Question 1:
Suppose this matrix `A` from activity 1 defines a linear inverse problem for an unknown vector $\mathbf{x}$ and an observed vector,
```{r}
b <- zeros(3,1)
b
```

Is there any vector $\mathbf{x}$ that satisfies,
$$\mathbf{A} x = \mathbf{b}?$$
Why does such a choice exist or why not?


#### Question 2:
Set a random seed to be equal to zero and generate a 4-by-4 dimension matrix with entries given by a sample size $n=16$ from the standard normal.  Suppose that this defines the relationship between the observed variable
```{r}
b <- ones(4,1)
```
and an unknown vector $\mathbf{x}$ as,
$$\mathbf{A}\mathbf{x}= \mathbf{b}.$$

Does any `x` as above exist?  If one does find the solution.  Explain why the solution exists or not due to the eigenvalues of the matrix.


