# Activity 10/21/2020

## STAT 445 / 645 -- Section 1001 <br> Instructor: Colin Grudzien<br>

## Instructions:
We will work through the following series of activities as a group and hold small group work and discussions in Zoom Breakout Rooms.  Follow the instructions in each sub-section when the instructor assigns you to a breakout room.

## Activities:

To perform some integration techniques, we will want the following packages,
```{r}
require(cubature)
require(caTools)
```

### Activity 1: Integration of density functions.

Let $X$ be a random variable distributed according to a CDF $F$, defining the probability distribution.  If $f$ is a continuous probability density function for that distribution, then
$$\int_{-\infty}^x f(t)\mathrm{d}t = F(x) = P(X \leq x)$$
We will now show how this relationship can be computed in R with the integrate command.

#### Question 1:
Recalling the form of the density function and CDF for the normal, find out by how much the integrate function applied to the density differs from the built-in CDF function for calculating $P(X \leq 1)$ for $X\sim N(0,1)$.  Call the `abs.error` variable to find the estimated error of the integral.  Is the estimate too high or too low in this example?


#### Question 2:
Recalculate the above, but set the `rel.tol` keyword in the `integrate` function to `10e-10`.  Now what is the result of the absolute difference with the `pnorm` calculation?  How does the error estimate look in relationship to the difference with the `pnorm` value?


#### Question 3:

We now change the standard normal to the normal with $\mu=-10$ and $\sigma=5$.  You must pass these arguments as keyword arguments in the `integrate` function, using the same keywords from the `dnorm` function.  The `integrate` will pass these arguments to the integrand function if specified correctly.  Try computing the integral of the density function for $N(-10,5)$ to compute $P(X\leq 1)$ using the `integrate` function with the default `rel.tol` argument and the `rel.tol=10e-10`.  How do these approximate integrals compare to the value from the `pnorm` function?


### Activity 2: Accuracy of integration by partition size

Consider how the accuracy of the trapezoidal rule differs from that of Gaussian quadrature.  In the following, we will define our ground truth for the true numerical value of the probability
$$\begin{align}
P(-2 \leq X \leq 2) & & X \sim N(0,1) 
\end{align}$$
to be
```{r}
ground_truth <- pnorm(2) - pnorm(-2)
```

The following `while` loop will approximate the integral 
$$\int_{-2}^2 f(x) \mathrm{d}x = P(-2 \leq X \leq 2) $$
with the trapezoidal rule, and it will double the step size until the error reaches the error level of the Gaussian quadrature
```{r eval=FALSE}
tol <-  # estimated error of Gaussian quadrature
```

Use the Gaussian quadrature method to integrate `dnorm()` over the interval $[-2,2]$ and assign the estimated error from the resulting object `abs.error` to the variable `tol` above the `while` loop  below.  Then find how large of a step size is needed to match the estimated error tolerance of the quadrature method.

```{r eval=FALSE}
absolute_error <-1 # dummy initial value for the error
n_part <- 2 # initial partition size

while (absolute_error > tol) {
   partition <- seq(-2, 2, len=n_part) # generate a partition
   value <- trapz(partition, dnorm(partition)) # compute the integral
   absolute_error <- abs(ground_truth - value) # compute the absolute error
   cat("Absolute error", absolute_error, "Partition size", n_part, "\n") # print the absolute error and partition size
   n_part <- n_part*2 # double partition size
}
```

