# Activity 09/17/2021

## STAT 445 / 645 -- Section 1001 <br> Instructor: Colin Grudzien<br>

## Instructions:
We will work through the following series of activities as a group.

## Activities:


### Activity 1: Set relationships

#### Exercise 1:
Suppose we are given the following sets.

```{r}
A <- c(1, 2, 3, 4)
B <- c(3, 4, 5, 6, 7)
```

Can you compute the following?

 1. $A\cap B$
 2. $A\cup B$
 3. $A \setminus B$

What is the probability of randomly selecting an odd number from each of the above sets?

##### Solution:

```{r}
intersect(A,B)
union(A,B)
setdiff(A,B)
```


The probability of selecting an odd number from $A\cap B = \{3,4\}$ is 1/2.

The probability of selecting an odd number from $A \setminus B = \{1,2\}$ is 1/2.

Respectively the probability of selecting an odd number from $A\cup B = \{1, 2, 3, 4, 5, 6, 7\}$ is 4/7.

#### Exercise 2:

Suppose that our total space of outcomes is given by $\Omega = \{1, 2, 3, 4, 5, 6, 7, 8, 9, 10\}$ and we have a random trial in which we select one of these numbers with equal probability.
Using our probability relationships, what is the probability of selecting a number in $A$ given that it is a number in $B$?

##### Solution:

We note that this can be represented by conditional probability as,
$$\begin{align}
P(A\vert B) &= \frac{P(A\cap B)}{P(B)} \\
&= \frac{2/10}{5/10} = \frac{2}{5}
\end{align}$$


### Activity 2: basic probability spaces

In the following you will need the `prob` package in R:

```{r, message=FALSE}
require(prob)
```

The following are standard experiments in the `prob` package that we use with the simple probability model.
```
 urnsamples(x, size, replace = FALSE, ordered = FALSE, ...),
 tosscoin(ncoins, makespace = FALSE),
 rolldie(ndies, nsides = 6, makespace = FALSE),
 cards(jokers = FALSE, makespace = FALSE),
 roulette(european = FALSE, makespace = FALSE).
```

In the following exercises, we will compute some complex conditional probabilities for these examples.

#### Exercise 1:

Suppose we roll 4 fair dies, compute the probability space as in the above and compute all the probabilities generally.

##### Solution:

```{r}
ev <- rolldie(times=4)
prob_space <- probspace(ev)
prob_space
```


#### Exercise 2:

From the above solution, compute the probability that the sum of all dice is greater than 20.

```{r}
Prob(prob_space, X1+X2+X3+X4 > 20)
```

#### Exercise 3:

Now compute the probability that the total of all dies is greater than 20, given that the first two rolls are greater than 4.

```{r}
Prob(prob_space,  event=X1+X2+X3+X4>20, given=(X1>4 & X2>4) )
```



