# Activity 09/09/2020

## STAT 445 / 645 -- Section 1001 <br> Instructor: Colin Grudzien<br>

## Instructions:
We will work through the following series of activities as a group and hold small group work and discussions in Zoom Breakout Rooms.  Follow the instructions in each sub-section when the instructor assigns you to a breakout room.

## Activities:

We will need both of the below for these activities.
```{r}
require(gapminder)
require(ggplot2)
```



### Activity 1: control flow



We will write a script that loops through the `gapminder` data by continent and prints out
whether the mean life expectancy for that continent is smaller or larger than 65 years in the year 1982.  
This will be walked through in several sub-problems.


#### Question 1:

We want to extract all the unique values of the continent vector.  We also need to loop over each of these continents and calculate the average life expectancy for each continent in 1982.


Try filling in the code below to simply loop over continents and print the average life expectancy over all years

```{r eval=FALSE}
for (continent_name in unique(gapminder$continent)){
  
  cat(continent_name, mean(continent_data$lifeExp), "\n")
}
```


##### Solution to question 1:

```{r}
for (continent_name in unique(gapminder$continent)){
  continent_data <- gapminder[gapminder$continent == continent_name, ]
  cat(continent_name, mean(continent_data$lifeExp), "\n")
}
```



#### Question 2:

Using the solution from the last example, fill in the code block to print the mean life expectancy of the continent in the year 1982.


##### Solution to question 2:


```{r eval=TRUE}
for (continent_name in unique(gapminder$continent)){
   continent_data <- gapminder[gapminder$continent == continent_name, ]
   continent_1982 <- continent_data[continent_data$year == 1982, ]
   cat(continent_name, mean(continent_1982$lifeExp), "\n")
}
```


#### Question 3:

We need to add an `if()` condition now before printing, which evaluates whether the calculated average life expectancy is above or below a threshold, and print an output conditional on the result.

Let's define the threshold
```{r}
threshold_value <- 65
```

Consider how to add an `if()` condition to the previous code to instead print  either
```{r eval=FALSE}
cat("Average Life Expectancy in", continent_name, "was less than or equal to", threshold_value, "in 1982.\n")
```

or 

```{r eval=FALSE}
cat("Average Life Expectancy in", continent_name, "was greater than", thresholdValue, "in 1982.\n")
```

##### Solution to question 3:

```{r eval=TRUE}
for (continent_name in unique(gapminder$continent)){
   continent_data <- gapminder[gapminder$continent == continent_name, ]
   continent_1982 <- continent_data[continent_data$year == 1982, ]
   continent_mean <- mean(continent_1982$lifeExp)
   if(continent_mean <= threshold_value){
       cat("Average Life Expectancy in", continent_name, "was less than or equal to", threshold_value, "in 1982.\n")
   }
   else{
       cat("Average Life Expectancy in", continent_name, "was greater than", threshold_value, "in 1982.\n")
        } # end if else condition
 
   } # end for loop

```


### Activity 2: plotting in ggplot2


Let's consider a plot below, where we use the `facet_wrap` with the `country` argument to plot the life expectancy versus year for all countries in the "Americas" region. 

```{r}
americas <- gapminder[gapminder$continent == "Americas",]
ggplot(data = americas, mapping = aes(x = year, y = lifeExp)) +
  geom_line() + facet_wrap( ~ country)
```

To clean this figure up for a publication we need to change some of the text elements.  The x-axis is too cluttered, and the y axis should read "Life expectancy", rather than the column name in the data frame.  We can do this by adding a couple of different layers. The **theme** layer controls the axis text, and overall text size. 



Labels for the axes, plot title and any legend can be set using the `labs` function. Legend titles
are set using the same names we used in the `aes` specification.  Thus below the color legend title is set using `color = "Continent"`, while the title
of a fill legend would be set using `fill = "MyTitle"`.

```{r}
ggplot(data = americas, mapping = aes(x = year, y = lifeExp, color=continent)) +
  geom_line() + facet_wrap( ~ country) +
  labs(
    x = "Year",              # x axis title
    y = "Life expectancy",   # y axis title
    title = "Figure 1",      # main title of figure
    color = "Continent"      # title of legend
  ) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


This is a taste of what you can do with ggplot2. RStudio provides a
really useful [cheat sheet][cheat] of the different layers available, and more
extensive documentation is available on the [ggplot2 website][ggplot-doc].
Finally, if you have no idea how to change something, a quick Google search will
usually send you to a relevant question and answer on Stack Overflow with reusable
code to modify!

[cheat]: https://www.rstudio.org/links/data_visualization_cheat_sheet
[ggplot-doc]: https://ggplot2.tidyverse.org/reference/


#### Question 1:

Try using the facet_wrap feature in ggplot2 to plot the life expectancies over time for all countries, but sorted by regions into sub-plots.  Use a scatter plot to picture life expectancies versus time, but where the points are colored by continent.  Finally, use the `geom_smooth` command to plot a linear model trend in each sub-plot.  The title should explain immediately the important information in the picture, with good, descriptive axis labels.  Use the `theme()` command to make sure the tick labels are also clear.


Note, if you want to save your plot to a file "regional_life_trends.png" you can use the command below where life_trend_plt is the plotting object assigned by the ggplot command.
```{r eval=FALSE}
# saving the figure to an output file, "regional_life_trends.png"
ggsave(filename = "regional_life_trends.png", 
       plot = life_trend_plt, width = 18, height = 15, 
       dpi = 300, units = "cm")
```



##### Solution to question 1:
```{r}
life_trend_plt <- ggplot(data=gapminder, 
                         aes(x = year, y = lifeExp, color = continent)) +
                  geom_point() +
                  geom_smooth(method = 'lm')  + facet_wrap(~continent) +
                  labs(
    x = "Year",                                # x axis title
    y = "Life expectancy",                     # y axis title
    title = "Regional life expectancy trends", # main title of figure
    color = "Region"                           # title of legend
  ) +
                  theme(axis.text.x = element_text(angle = 90, hjust = 1))

life_trend_plt
```
