# Activity 09/01/2021

## STAT 445 / 645 -- Section 1001 <br> Instructor: Colin Grudzien<br>

## Instructions:
We will work through the following series of activities as a group.  Follow the instructions in each sub-section when the instructor assigns you to a breakout room.

## Activities:

### Activity 1: vectors and matrices


#### Question 1:
We can declare a matrix full of zeros:

```{r eval=FALSE}
matrix_example <- matrix(0, ncol=6, nrow=3)
matrix_example
```

And similar to other data structures, we can ask things about our matrix:

```{r eval=FALSE}
class(matrix_example)
typeof(matrix_example)
str(matrix_example)
dim(matrix_example)
nrow(matrix_example)
ncol(matrix_example)
```

What do you think will be the result of
`length(matrix_example)`?
Try it.
Were you right? Why / why not?


#### Question 2:
Make matrix containing the numbers `1:50`, with 5 columns and 10 rows.
Did the `matrix` function fill your matrix by column, or by row, as its default behavior?
See if you can figure out how to change this.
(hint: read the documentation for `matrix`!)


### Activity 2: practice with subsetting data

#### Question 1:

Given the following code:
```{r}
x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
names(x) <- c('a', 'b', 'c', 'd', 'e')
print(x)
```

Come up with at least 2 different commands that will produce the following output:
##### Solution:
```{r}
x[2:4]
x[-c(1,5)]
x[c('b','c','d')]
```

#### Question 2:

Given the following code:

```{r eval=FALSE}
x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
names(x) <- c('a', 'b', 'c', 'd', 'e')
print(x)
```

Write a subsetting command to return the values in x that are greater than 4 and less than 7.

##### Solution

```{r}
x[ x > 4 & x <7]
```


#### Question 3:

Selecting elements of a vector that match any of a list of components
is a very common data analysis task. For example, the gapminder data set
contains `country` and `continent` variables, but no information between
these two scales. Suppose we want to pull out information from southeast
Asia: how do we set up an operation to produce a logical vector that
is `TRUE` for all of the countries in southeast Asia and `FALSE` otherwise?

Suppose you have these data:
```{r}
seAsia <- c("Myanmar","Thailand","Cambodia","Vietnam","Laos")
require(gapminder)
## extract the `country` column from a data frame (we'll see this later);
## convert from a factor to a character;
## and get just the non-repeated elements
countries <- unique(as.character(gapminder$country))
#print(countries)
```

There's a wrong way (using only `==`), which will give you a warning;
a clunky way (using the logical operators `==` and `|`); and
an elegant way (using `%in%`). See whether you can come up with all three
and explain how they (don't) work.

```{r}
# wrong
countries == seAsia
# clunky
countries == "Myanmar" | countries == "Thailand" | countries == "Cambodia" | countries == "Vietnam" | countries == "Laos"
# elegant
countries %in% seAsia
```


To extract a dataframe associated to Southeast Asia, we can write

```{r}
gapminder[as.character(gapminder$country) %in% seAsia,]
```

#### Question 4:

Consider the R output of the matrix below:
```{r}
matrix(c(4, 1, 9, 5, 10, 7), ncol = 2, byrow = TRUE)
```
Examine each command below and try to figure out the output before typing them.

  1. `matrix(c(4, 1, 9, 5, 10, 7), nrow = 3)`
  2. `matrix(c(4, 9, 10, 1, 5, 7), ncol = 2, byrow = TRUE)`
  3. `matrix(c(4, 9, 10, 1, 5, 7), nrow = 2)`
  4. `matrix(c(4, 1, 9, 5, 10, 7), ncol = 2, byrow = TRUE)`
  
