# Activity 10/18/2021

## STAT 445 / 645 -- Section 1001 <br> Instructor: Colin Grudzien<br>

## Instructions:
We will work through the following series of activities as a group and hold small group work and discussions in Zoom Breakout Rooms.  Follow the instructions in each sub-section when the instructor assigns you to a breakout room.

## Activities:

To perform some matrix operations, it can be helpful to use some non-base functions designed for this.  We will require the library `matlab`

```{r}
require("matlab")
```


### Activity 1: Diagonalization

Suppose we are given the following matrix
```{r}
A <- matrix(seq(3, 27, by=3), nrow=3, ncol=3)
```

#### Question 1:
Compute the rank of the matrix `A` . Then compute the eigen values of the matrix `A`.  How is the rank related to the eigenvalues of the matrix?

##### Solution:

We compute the rank as follows:
```{r}
qr(A)$rank
```

The eigen decomposition is computed as:
```{r}
eigen(A)
```

We see that the rank of the matrix `A` is 2, and that it has exactly 2 non-zero eigenvalues.

#### Question 2:
Use the  eigen decomposition as follows: extract the matrix of the eigenvectors by calling the variable `$vectors` and assign these vectors to a matrix called `C`.  Check if it is possible to invert this matrix and if so, define the inverse `C_inv`.  Explain why this is possible or why not.


##### Solution:

We define the eigen decompostion and the matrix `C` as follows.
```{r}
e_decomp <- eigen(A)
C <- e_decomp$vectors
```

Notice,
```{r}
det(C)
```
so that it is invertible.  We may look instead at
```{r}
eigen(C)
```
to notice that this has complex, non-zero eigenvalues.

Therefore, we assign `C_inv` as follows:
```{r}
C_inv <- solve(C)
```


#### Question 3: 
Now try the following, multiply `A` by `C_inv` on the left and `C` on the right.  What do you notice about the values off diagonal? 

Try using the following comparison to see how close the elements are to the value zero,
```{r, eval=FALSE}
abs(product) <= ones(3) * 10e-14
```

Take the diagonal elements of the product above and find the absolute difference of these from the eigenvalues of `A`. What do you notice about the result.

##### Solution:

```{r}
product <- C_inv %*% A %*% C
product
abs(product) <= ones(3) * 10e-14
```
This is numerically a diagonal matrix, with a zero also in the last diagonal position.

If we take the absolute difference of the diagonal with the eigenvalues, we get
```{r}
abs(diag(product) - e_decomp$values)
```
We see that these are indeed the eigenvalues.  Multiplying on the left and right as above gave a change of coordinates to the diagonal form of the transformation represented by `A`.

