# Activity 11/08/2021

## STAT 445 / 645 -- Section 1001 <br> Instructor: Colin Grudzien<br>

## Instructions:
We will work through the following series of activities as a group and hold small group work and discussions in Zoom Breakout Rooms.  Follow the instructions in each sub-section when the instructor assigns you to a breakout room.

## Activities:

We will use the following package for the multivariate normal distribution:
```{r}
require(mnormt)
```

### Activity 1: The multivariate normal and the Mahalanobis transformation

#### Question 1:
Create a 2x2 matrix $\mathbf{A}$ with the columns given by the integers 1 through 4 in order.  Evaluate if this has any zero eigenvalues.  Then construct the symmetric matrix by $\mathbf{B} =\mathbf{A}\mathbf{A}^\top$.  Evaluate if this has any zero eigenvalues.

##### Solution:

```{r}
A <- matrix(1:4, 2, 2)
eigen_decomp_A <-eigen(A)
eigen_decomp_A
B <- A %*% t(A)
eigen_decomp_B <- eigen(B)
eigen_decomp_B
```

#### Question 2:

Use the following seed and the `rmnorm` function to generate a random sample.   Set the mean equal to mu below, but try each of $\mathbf{A}$ and $\mathbf{B}$ for the `varcov` parameter.  What do you notice?  Plot the sample and the lines formed by the eigenvectors for the `varcov` parameter using the supplied plotting code below. What do you notice about the plot?
```{r}
set.seed(123)
sample_size <- 10e3
mu = matrix(34:35, 2)
```

```{r, eval=FALSE}
eigen_decomp <-
vecs <- 
vals <- 
eigScl  <- vecs %*% diag(sqrt(vals))  # scale eigenvectors to length = square-root of eigenvalue
xMat    <- rbind(mu[1] + eigScl[1, ], mu[1] - eigScl[1, ])
yMat    <- rbind(mu[2] + eigScl[2, ], mu[2] - eigScl[2, ])
matlines(xMat, yMat, lty=1, lwd=2, col="red")
```

##### Solution:
We can only supply a positive (semi-)definite matrix as a covariance by definition. `A` is not symmetric so this does not satifsy the condition, but `B` is symmetric by construction.

```{r}
mv_norm_sample <- rmnorm(n=sample_size, mean=mu, varcov=B)
plot(mv_norm_sample[,1], mv_norm_sample[,2], asp=1)
eigen_decomp <- eigen(B)
vecs <- eigen_decomp$vectors
vals <- eigen_decomp$values
eigScl  <- vecs %*% diag(sqrt(vals))  # scale eigenvectors to length = square-root of eigenvalue
xMat    <- rbind(mu[1] + eigScl[1, ], mu[1] - eigScl[1, ])
yMat    <- rbind(mu[2] + eigScl[2, ], mu[2] - eigScl[2, ])
matlines(xMat, yMat, lty=1, lwd=2, col="red")
```


#### Question 3:
Now try subtracting the value `mu` from the sample in all entries.  Compute the mean of the sample.  Plot the transformed sample, and the eigen vectors as before with the code below:

```{r, eval=FALSE}

plot(transformed_sample[,1], transformed_sample[,2], asp=1)
xMat    <- rbind(0 + eigScl[1, ], 0 - eigScl[1, ])
yMat    <- rbind(0+ eigScl[2, ], 0 - eigScl[2, ])
matlines(xMat, yMat, lty=1, lwd=2, col="red")
```

##### Solution:

```{r}
transformed_sample <- mv_norm_sample - matrix(mu, sample_size, 2, byrow=TRUE)
colMeans(transformed_sample)
plot(transformed_sample[,1], transformed_sample[,2], asp=1)
xMat    <- rbind(0 + eigScl[1, ], 0 - eigScl[1, ])
yMat    <- rbind(0+ eigScl[2, ], 0 - eigScl[2, ])
matlines(xMat, yMat, lty=1, lwd=2, col="red")
```


#### Question 4:
Now, try transforming the sample by the inverse of `B` and the inverse of `A`.  Plot the results and see what you get.  What do you notice about the results?


##### Solution:

In this set up, `A` is playing the role in some sense of the standard deviation in which we can standardize a sample by subtracting the mean and dividing by the square root of the variance.  Dividing by `B` does not give the desired result.

```{r}
transformed_sample_B <-  solve(B) %*% t(transformed_sample)
transformed_sample_B <- t(transformed_sample_B)
plot(transformed_sample_B[,1], transformed_sample_B[,2], asp=1)
```

However, dividing by `A` gives a sample according to the standard multivariate normal:

```{r}
transformed_sample_A <-  solve(A) %*% t(transformed_sample)
transformed_sample_A <- t(transformed_sample_A)
plot(transformed_sample_A[,1], transformed_sample_A[,2], asp=1)
```

This is known as the Mahalanobis transform.