# Activity 09/29/2021

## STAT 445 / 645 -- Section 1001 <br> Instructor: Colin Grudzien<br>

## Instructions:
We will work through the following series of activities as a group and hold small group work and discussions in Zoom Breakout Rooms.  Follow the instructions in each sub-section when the instructor assigns you to a breakout room.

## Activities:

### Activity 1: Computing z-scores and critical values

#### Question 1:

Recall, a general normal confidence interval (with known variance $\sigma^2$) is given of the form:

$$\left(\overline{x} - z_\frac{\alpha}{2}\frac{\sigma}{\sqrt{n}}, \overline{x}+z_\frac{\alpha}{2} \frac{\sigma}{\sqrt{n}}\right)$$
where $z_\frac{\alpha}{2}$ is selected so that 

$$P\left(Z\leq z_\frac{\alpha}{2}\right) = 1 - \frac{\alpha}{2}$$.

The confidence level for the above interval is thus equal to $(1-\alpha)\times 100\%$.

For a normal population with known variance $\sigma^2$, what is the confidence level (written as a percentage) for the interval
$$\left(\overline{x} - 2.14\frac{\sigma}{\sqrt{n}}, \overline{x}+2.14 \frac{\sigma}{\sqrt{n}}\right)?$$

##### Solution:

We notice that $z_\frac{\alpha}{2} = 2.14$ and that $P(Z\leq 2.14)$ is given as
```{r}
pnorm(2.14)
```
Therefore, we consider that
$$\begin{align}
&P(Z\leq z_\frac{\alpha}{2}) = 1 - \frac{\alpha}{2}\\
\Leftrightarrow & \alpha = 2 - 2 P(Z \leq z_\frac{\alpha}{2})
\end{align}$$
```{r}
alpha <- 2- 2*pnorm(2.14)
alpha
```

The confidence level is then given as

```{r}
(1 - alpha)*100
```
percent.

### Activity 2: Computing point estimates and confidence intervals

#### Question 1:

The yield of a chemical process is being studied. From previous experience, yield is known to be normally distributed and $\sigma=3$. The past five days of plant operation have resulted in the yields in the following array:
```{r}
my_data <- c(91.6, 88.75, 90.8, 89.95, 91.3)
```

Compute each of the:

1. sample mean
2. sample standard deviation
3. the standard error of the sample mean
4. the sample-estimate for the standard error of the sample mean
5. the 95% two-sided confidence interval for the true mean
6. the 99% two-sided confidence interval for the true mean

##### Solution

1. Notice, we can compute the sample mean as

```{r}
sample_mean <- mean(my_data)
sample_mean
```

2. The sample standard deviation is given as

```{r}
sample_std <- sd(my_data)
sample_std
```

Notice, it does not equal the population standard deviation -- this is sampling error.

3. The standard error of the sample mean with the known variance is given as

$$\frac{\sigma}{\sqrt{n}} = \frac{3}{\sqrt{5}}$$ 
or
```{r}
SE <- 3/sqrt(length(my_data))
```

4. If the true standard deviation $\sigma$ is unknown, we estimate it with the sample standard deviation,

```{r}
sample_std
```

The sample estimate for the standard error is thus equal to 
$$\frac{s}{\sqrt{n}}$$
or
```{r}
SE_estimated <- sample_std / sqrt(length(my_data))
SE_estimated
```

5. With the sample mean and the standard error, we can compute confidence interval as

$$\left(\overline{x} - z_\frac{\alpha}{2}\frac{\sigma}{\sqrt{n}}, \overline{x}+z_\frac{\alpha}{2} \frac{\sigma}{\sqrt{n}}\right)$$

using the appropriate critical value.

For the 95% confidence interval, we need to find the value for which $1-\frac{\alpha}{2} = 1-0.025=0.975$ of the area lies to the left and $\frac{\alpha}{2}=0.025$ lies to the right.

```{r}
z_alpha_over_2 <- qnorm(.975)
CI_95 <- c(sample_mean - SE * z_alpha_over_2, sample_mean + SE * z_alpha_over_2)
CI_95
```

6. Likewise, we can find

$$1-\frac{\alpha}{2}=1-0.005=0.995$$

Therefore,
```{r}
z_alpha_over_2 <- qnorm(.995)
CI_99 <- c(sample_mean - SE * z_alpha_over_2, sample_mean + SE * z_alpha_over_2)
CI_99
```

### Activity 3: Central limit theorem

#### Question 1:
Recall, we can generate a random sample of the Poisson distribution with the `rpois` function.  Let us set $\lambda=25$ and draw a sample of size `n=20` from this distribution, and compute the sample mean.  Let's replicate the process and plot the sample means over $m=50$ replications. Plot the corresponding central limit theorem sampling distribution for the mean along with these values, with a range given over $[20, 30]$.

```{r}
set.seed(0)
lambda <- 25
m <- 50
n <- 20
sigma <- sqrt(lambda)
se <- sigma / sqrt(n)
f <- dnorm(seq(20, 30, by=0.01), mean=25, sd=se)
plot(seq(20, 30, by=0.01), f, type = "l", main = "Sampling distribution for sample mean", xlab = "x", ylab = "Density")
for (i in 1:m){
 sample <- rpois(n, lambda = lambda)
 sample_mean <- mean(sample)
 points(sample_mean, 0, col="red")
}
```

